Experience Report: Haskell in Computational Biology

### 2012-02-15 mrfy_experience_report.tex
Lots of outlining, included files, math. Little new written content.

\usepackage{amsmath}
\usepackage{listings}
\lstset{
    tabsize = 2
    }

@@

[insert Viterbi recurrence relations]

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container (again, a 
three-dimensional array). We determined that the run-time performances of both 
approaches were virtually indistinguishable, and we also noted that the 
top-down version quite faithfully resembled the recurrence relations found in a 
textbook~\cite{durbin} or the HMMER literature~\cite{eddy}. We chose to keep 
the top-down version, which later proved to be wise when we had to track down a 
bug in one of our base cases. This bug was observed as a missing state in the 
alignment our program displayed, and we quickly discovered that one of our base 
cases was missing this state: it was returning an empty list as the base of the 
path, rather than a list containing a single node. In this, we were grateful 
for the resemblance between the mathematical description of the algorithm and 
the top-down dynamic programming approach in Haskell, which resulted in 
perspicuous code.

[part of Viterbi code]
\begin{figure*}
\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
\end{figure*}

@@

We wrote a basic search function in only twenty-two lines of code; as one of 
its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques. We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies. For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions. In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}. We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.

[listing for search function, and constructor for SearchStrategy]

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type \texttt{type 
ScorePathCons a = a -> [a] -> [a]} and two functions conforming to that type. 
Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
or \texttt{consNoPath} as a parameter, the path would or would not be 
allocated. We found this higher-order function approach to be simple in code 
and in concept, and it imposes negligible run-time overhead as compared to 
having two distinct functions.

@@

\section{Barriers}

- profiling
- debugging

>>>>

\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

We implemented the Viterbi algorithm in two distinct ways. We first wrote a 
bottom-up dynamic programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model. The array containing the paths 
served as the memoization table. We next wrote a top-down dynamic programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both). In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array). We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
(see Equation \ref{viterbi_eqn}). We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases. This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node. In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down dynamic 
programming approach in Haskell, which resulted in perspicuous code.

\begin{figure*}
\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
\end{figure*}

@@

We wrote a basic search function in only twenty-two lines of code; as one of 
its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques. We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies. For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions. In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}. We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant. In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this. We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

\begin{figure*}
\lstinputlisting[label=search,caption=Search]{search_simple.hs}
\end{figure*}

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type. 
Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
or \texttt{consNoPath} as a parameter, the path would or would not be 
allocated. We found this higher-order function approach to be simple in code 
and in concept, and it imposes negligible run-time overhead as compared to 
having two distinct functions.

@@

\section{Our previous code base compared}

 - SMURF bad
 - HMMER good

\section{State of the practice}

\subsection{Our group then}
 - Modifying Formatt took nearly a year
 - mutable data structures hard to repurpose
 - one enhancement deemed infeasible
 - C++ failures most commonly segfaults, memory errors, heap exhaustion

\subsection{Our group now}
 - Higher order functions militate towards more flexibility
 - Haskell failures are usually typecheck failures; at worst runtime errors around bounds checking
 - MRFy about 3 months of part-time work

\subsection{Computational biology at large}
 - some good software, like HMMER

\section{What we think about it}

\subsection{Barriers}

- profiling
- debugging
 [ talk about profiling and debugging problems where?]
[ something about quickcheck: how do we know when we have enough QC props? How do we start testing a complex module? ]

\subsection{What you must know to succeed}
- Ecosystem is loaded with tools that look promising but either are no longer maintained or are not ready for prime time (Pads, Backtrace)

\subsection{Evaluation}
- What does the local expert do? Do you have to have one?

- To the FP community: Those who want us to abandon our old tools for debugging imperative code have not provided a clear path to follow (both intellectually, and via software tools).

- No obvious body of knowledge on code improvement.

- The ecosystem contains tools that look promising and work great: Data.Memocombinators, Quickcheck, Parallel Strategies. How to tell what's good? How did we discover what's good?
- Profiler can produce backtraces, but this was hard to discover.
